---
title: "Dice Rolling"
---

```{r, include = F}
library(dndR)
```

At its simplest, DnD involves significant amounts of dice rolling and (often) summing their values. So, `dndR` includes a `roll` function! This function supports 'rolling' up to 10 million of any of the standard dice and summing their results. "Standard" dice include the following numbers of sides: 100, 20, 12, 10, 8, 6, 4, and 2.

Dice names are abbreviated as the <u>number</u> of dice to roll], a <u>lowercase "d"</u>, and the <u>type</u> of dice to roll

For instance, to roll 3 six-sided dice we could do the following:
```{r roll_fxn_1}
dndR::roll(dice = '3d6')
```

The `roll` function returns the total of the side of the dice as a number so if you want to add multiple rolls together you can treat the function calls themselves as if they are numbers!

```{r roll_fxn_2}
dndR::roll('3d8') + dndR::roll('1d4')
```

Also, rolling two 20-sided dice (i.e., "2d20") is automatically assumed to be rolling with either advantage or disadvantage so it will return a message to that effect and both numbers.

```{r roll_fxn_3}
dndR::roll(dice = '2d20')
```

