---
title: "Functions For GMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions For GMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libs, echo = F, message = F}
# devtools::install_github("njlyon0/dndR", force = TRUE)
library(dndR)
```

## Balancing Encounters

When I am the Dungeon/Game Master (DM / GM) I find encounter balancing to be really difficult, in part because of the nest of interrelated tables and experience point multipliers outlined in the Dungeon Master's Guide (DMG) that must be consulted for novice GMs such as myself. To help other newbies, `dndR` includes two core functions aimed at simplifying encounter balancing.

### Experience Point (XP) Thresholds

The difficulty of an encounter in D&D is affected by three things:

1. Level of characters
2. Number of characters in the party
3. How difficult the GM wants to make things for their players

The DMG handles this by providing experience point (XP) thresholds based on these three factors. All enemies are worth a pre-determined amount of XP so encounters are balanced by the DMG listing the total XP of all monsters in a given fight for every level of players, party size, and difficulty. That table is useful but a little dense to work through as you're prepping potentially multiple encounters per session, so this is where `xp_pool` becomes useful.

`xp_pool` returns the amount of XP the GM can 'spend' on monsters in a given encounter to ensure the difficulty is as desired based on the three factors identified above.

```{r xp_pool_fxn}
dndR::xp_pool(party_level = 2, party_size = 4, difficulty = "medium")
```

In the above case, the DM can have their four-person, level two party face creatures worth a total of 625 XP and be reasonably confident that the difficulty will be somewhere in the middle.

### XP Multipliers

However, while knowing the amount of XP available to the DM per encounter is helpful, it fails to account for the effect of the *number* of enemies. A fight versus a single monster worth 1000 XP is a very different proposition than a fight against four creatures each worth 250 XP even though the total XP is the same.

The DMG accounts for this by providing XP multipliers based on the number of monsters and the number of players. The same total monster XP is multiplied by a larger value for more monsters facing a smaller party than it would be for fewer monsters facing a larger party.

So, if you are using the DMG to balance an encounter you have to total up the XP of the monsters in an encounter and then go to a different part of the DMG where you can multiply that value by the multiplier in the relevant row of a second table which would then tell you the "actual" XP of the creatures you have selected. If you had too many or too few monsters you'd have to repeat this process iteratively until you identified the correct amount of "raw" XP to yield your desired "realized" XP in an encounter. Cumbersome, right?

`dndR`'s `xp_cost` function provides an alternative to using the multiplier table. `xp_cost` requires the "raw" XP of the monsters you have selected, the number of monsters, and the number of player characters (PCs) and returns the "realized" XP. You can then quickly compare this with the value return by `xp_pool` to determine whether you need to add or remove creatures from your encounter.

```{r xp_cost_fxn}
dndR::xp_cost(monster_xp = 1000, monster_count = 2, party_size = 3)
```

### Balancing Demonstration

That was a lot of content so let's run through a quick example to see how `xp_pool` and `xp_cost` work together to make encounter balancing simpler. Let's say I am running a game for four players, all level 3, and I want to design a hard encounter for them.

To begin, I'd identify the total XP I can spend to make an encounter of this difficulty using `xp_pool`.

```{r xp_demonstration_1}
dndR::xp_pool(party_level = 3, party_size = 4, difficulty = 'hard')
```

Okay, so the pool of XP available for me to spend is 1,112. Keeping that threshold in mind, I can use `xp_cost` to check the value of two monsters worth (total) 500 XP against that threshold.

```{r xp_demonstration_2}
dndR::xp_cost(monster_xp = 500, monster_count = 2, party_size = 4)
```

I can see that I'm well under the XP threshold I have to play with so I can add a monster and see where that leaves me.

```{r xp_demonstration_3}
dndR::xp_cost(monster_xp = 750, monster_count = 3, party_size = 4)
```

A little over because of the multiplier, so let's say I change my mind and keep three monsters but change their identity to one with a lower XP value.

```{r xp_demonstration_4}
dndR::xp_cost(monster_xp = 600, monster_count = 3, party_size = 4)
```

Basically right on target! I can now pick out my three monsters that total up to 600 XP raw and know that they will likely* make a hard encounter for my players! (* "Likely" because there is dice rolling involved and it is possible that the monsters roll well while my players roll badly or vice versa).

## Monster Creation

Creatures that you create (or "homebrew") can be a great way to add flavor to an encounter or can even form the centerpiece of a larger campaign arc! `dndR`'s `monster_stats` and `monster_creator` functions can help DMs to quickly stat out their custom monsters

### Finding Official Monster Statistics

The DMG provides a table (see p. 274) that gives the vital statistics of creatures based on their Challenge Rating (CR) but this table can be cumbersome to compare to Experience Points (you know, the things used to determine how hard an encounter will be for your party?).

`monster_stats` streamlines this process by allowing you to either

1. Input the XP you want to spend on this creature (you can use the value returned by `xp_cost`) 

*OR*

2. Input the challenge rating (CR) if you know it.

Once either XP or CR is provided, `monster_stats` returns the creature's statistics as they appear in the DMG for a *single* creature of that difficulty.

```{r monster_stats_fxn1}
dndR::monster_stats(cr = 7)
```

Challenge Rating is more than a little esoteric so feel free to ignore that argument entirely if XP is more comfortable for you! Note that XP is rounded down to the nearest specified monster "tier" specified in the DMG.

```{r monster_stats_fxn2}
dndR::monster_stats(xp = 8000)
```

If you specify both CR and XP, the function prints a warning and proceeds *only* with CR (i.e., ignoring whatever you put in as XP). I figured if a user is confident enough to put CR at all it should be given priority but am happy to tweak this if desired!

```{r monster_stats_fxn3}
dndR::monster_stats(xp = 1000, cr = 4)
```

### Making Custom Monsters

If you'd rather take a more individualized approach, you can use `monster_creator` instead. This function follows the advice of [Zee Bashew](https://twitter.com/Zeebashew) on how to build interesting, challenging monsters for your party. These monsters are built somewhat according to the Dungeon Master's Guide for creating monsters, partly Zee's [YouTube video on homebrewing monsters based on the videogame *The Witcher*](https://www.youtube.com/watch?v=GhjkPv4qo5w), and partly on my own intuition about scaling the difficulty of a creature. 

Creatures made by `monster_creator` have the following information in addition to the statistics returned by `monster_stats`:

- Up to **five** damage *resistances*
- Up to **two** damage *immunities*
- **One** *vulnerability*
- Proficiency in **two** saving throws
- Fully custom ability scores (returned only as their modifiers)

This combination allows you to build complex and mysterious homebrew monsters with plenty of opportunities for the party to have to investigate around to discover the monster's strengths and weaknesses before the final showdown.

```{r monster_creator}
dndR::monster_creator(party_level = 5, party_size = 4)
```

Creature's special traits are chosen randomly so you may want to re-run the function several times (or mentally modify one or more parts of the output) to get a monster that fits your campaign and players. 

Note that if you use `monster_creator` you may want to help your players identify the creature's immunities and vulnerabilities *before* the actual confrontation with the creature to avoid sending them into a fight that is more difficult than your party can handle.

## Non-Player Characters (NPCs)

> **DM:** "You walk into the tavern and glance around the poorly lit room. In a corner booth, you catch sight of the betrayer you've been hunting these past two weeks--Tymoth Grissel. You nearly had him in the Sootpile mountains but that rockslide allowed him to escape by the skin of his teeth. He won't escape this time."

> **Player:** "Cool, cool...but who *else* is in the tavern?

> **DM:** ...

When I'm DMing, I sometimes struggle with improvising background NPCs (non-player characters) that aren't relevant to the current questlines when my player ask about them. To hopefully help fellow DMs, `dndR` includes a function to let you quickly respond in situations like the one I've described above.

`npc_creator` quickly selects a random race and job for however many NPCs you need. The function returns this as a dataframe in case you had some follow-up processing you wished to do but I find myself running the function without assigning it to an object so that it prints in the console.

```{r npc-creator-fxn}
dndR::npc_creator(npc_count = 3)
```

If your players decide to go chat with the goblin carpenter passing them on the street (or whoever else they cross paths with) you can then improv more details about the NPC as needed. Hopefully this function helps you skip the "uh..." step of trying to on-the-fly come up with some brief NPC descriptions.

Note that because both NPC race and job are selected randomly you may wind up with some unusual combinations. I hope that adds to the fun but feel free to re-run `npc_creator` until you get NPCs that fit your world and intended tone.

## Party Diagram

When designing encounters (or even a longer campaign arc) it can be helpful to know your party's strengths and weaknesses. If your party on average has great CON scores, it is nice to know that when deciding the kinds of monsters and plot lines you'll weave in. This isn't to say that you'd identify your party's weaknesses and then focus only on those (or at least not *only* that). You could just as easily play to your party's strengths and help set the stage for miraculous successes that you know are statistically probable. If this sounds valuable to you, use `party_diagram` to generate a plot of your player's ability scores either by player or by ability score. You can either provide the PC's ability scores interactively in response to prompts from the function or you can provide a named list of scores per PC.

Also, I want to briefly give a huge shout out to [Tim Schatto-Eckrodt](https://kudusch.de/) for contributing this function!

### Preparing Party Scores

First, I'll define the party in this example because I can't interactively respond to the prompts in this website build. Let's say I have a three player party consisting of Vax, Beldra, and Rook.

```{r diagram_define}
party_list <- list(
  Vax = list(STR = "10", DEX = "13", CON = "14",
             INT = "15", WIS = "16", CHA = "12"),
  Beldra = list(STR = "20", DEX = "15", CON = "10",
                INT = "10", WIS = "11", CHA = "12"),
  Rook = list(STR = "10", DEX = "10", CON = "18",
              INT = "9", WIS = "11", CHA = "16")
   )
```

### Party Diagram - by Player

Now that we have the list of player statistics, we can give it to `party_diagram` to check out our scores by player.

```{r diagram_by-player}
dndR::party_diagram(by = "player", pc_stats = party_list, quiet = TRUE)
```

### Party Diagram - by Ability

If you want to visualize by ability score instead, you only have to change the `by` argument!

```{r diagram_by-ability}
dndR::party_diagram(by = "ability", pc_stats = party_list, quiet = FALSE)
```

The `quiet` argument is useful when you entered your party's statistics interactively as it returns the simple R call you'd need to assemble it for future use. I recommend copy/pasting that output into an R script or note for later if you think it will be valuable.

