---
title: "Dice Rolling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dice Rolling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libs, echo = F, message = F}
# devtools::install_github("njlyon0/dndR", force = TRUE)
library(dplyr); library(ggplot2); library(magrittr); library(tidyr)
```

## Welcome to `dndR`!

I am a big fan of both R and Dungeons and Dragons so I thought it would be fun to combine the two and build an R package that supports other fans of this hobby! `dndR` includes functions for several facets of D&D and I've broken these categories into separate, dedicated vignettes. As dice rolling is arguably the most fundamental element of D&D, this first vignette focuses on the dice rolling functions included in the package.

If any of these functions break for you, please [post an Issue](https://github.com/njlyon0/dndR/issues) so that I can fix the issue as soon as possible.

## Dice Rolling

At its simplest, D&D involves significant amounts of dice rolling and (often) summing their values, so `dndR` includes a `roll` function! This function supports 'rolling' up to 10 million of any of the standard dice and summing their results. "Standard" dice include the following numbers of sides: 100, 20, 12, 10, 8, 6, 4, and 2.

```{r roll_fxn_1}
dndR::roll('3d6')
dndR::roll('3d8') + dndR::roll('1d4')
```

Note that '2d20' is automatically assumed to be rolling with advantage/disadvantage so it will return a message to that effect and both numbers.

```{r roll_fxn_2}
dndR::roll(dice = '2d20')
```

You can also use `probability_plot` to roll dice a specified number of times and generate a `ggplot2` graph of the frequency of various outcomes. The median outcome is specified by a dashed vertical line.

```{r prob_plot_fxn, warning = F, error = F, fig.height = 3, fig.width = 6, fig.align = "center"}
dndR::probability_plot(dice = "2d6", roll_num = 499)
```

## Character Creation

`pc_creator` rolls for a character's ability scores (strength, dexterity, constitution, intelligence, wisdom, and charisma) given a particular class, race, and preferred method of rolling for ability scores.

```{r pc_creator_fxn}
dndR::pc_creator(class = 'barbarian', race = 'half orc', score_method = "4d6")
```

You can check which classes and races are currently supported by `pc_creator` by running `dnd_classes` or `dnd_races`. If you have a class/race in mind that isn't supported you can [post an Issue](https://github.com/njlyon0/dndR/issues) and I'll add that class/race's stats to the function ASAP!

While waiting for me to act on your Issue, you can run the simpler `ability_scores` function to simply roll for ability scores and manually assign them to specific abilities and handle race/class/background based modifiers yourself.

```{r ability_scores_fxn}
dndR::ability_scores(method = "4d6")
```

If you do experience point-based leveling, you can use the `pc_level_calc` function to check what level your PC has reached with the amount of XP they have earned. Thanks to Humberto Nappo for contributing this function!

```{r pc_level_calc}
dndR::pc_level_calc(player_xp = 4100)
```

## Spell-Related Functions

There are _a lot_ of spells in the various Dungeons & Dragons source books that your character might be able to invoke. Finding specific spells can be a daunting task in the face of the sheer number of spells that exist so `dndR` has some functions to hopefully make this effort more manageable for your magic-using characters and NPCs.

`spell_list` lets you search through official spells based on several categories. A dataframe including all of the big picture information for spells that meet your criteria is returned to help orient you for any subsequent steps you plan on taking. Note that all `spell_list` arguments that accept characters are case insensitive and queries use partial string matching so you don't need to try to engineer exact matches. Specifically, you can query spells based on one or more of the following criteria:

- Spell name(s)
- Character classes with access to the spell(s)
- Minimum spell slot required to cast the spell and/or "cantrip"
- School(s) of magic to which the spell belongs (e.g., "abjuration", etc.)
- Whether or not the spell can be cast as a ritual (note this argument accepts only `TRUE` or `FALSE`)
- Time to cast the spell (either the phase of a turn required or the in-game time for spells that take more than one turn to cast)

```{r spell_list_fxn}
dndR::spell_list(name = "bolt", class = "sorcerer", ritual = FALSE)
```

If you have a specific spell (or several spells) for which you want the complete spell information (including the description text), `spell_text` is there to help. `spell_text` only accepts the spell name(s) for which you want information and returns all of the information that `spell_list` does but also gives you the spell description text and (if applicable) how the spell changes when cast with a higher level spell slot or by a higher level player character.

```{r spell_text_fxn}
dndR::spell_text(name = "chill touch")
```

## Encounter Balancing

When I am the Dungeon/Game Master (DM / GM) I find encounter balancing to be really difficult, in part because of the nest of inter-related tables and experience point multipliers outlined in the Dungeon Master's Guide (DMG) that must be consulted for novice GMs such as myself. To help other newbies, `dndR` includes the `xp_pool` and `xp_cost` functions.

### Experience Point (XP) Thresholds

The difficulty of an encounter in DnD is affected by three things:

1. Level of characters
2. Number of characters in the party
3. How difficult the GM wants to make things for their players

The DMG handles this by providing experience point (XP) thresholds based on these three factors. All enemies are worth a pre-determined amount of XP so encounters are balanced by the DMG listing the total XP of all monsters in a given fight for every level of players, party size, and difficulty. That table is useful but a little dense to work through as you're prepping potentially multiple encounters per session, so this is where `xp_pool` becomes useful.

`xp_pool` returns the amount of XP the GM can 'spend' on monsters in a given encounter to ensure the difficulty is as desired based on the three factors identified above.

```{r xp_pool_fxn}
dndR::xp_pool(party_level = 2, party_size = 4, difficulty = "medium")
```

### XP Multipliers

While it is crucial to know the amount of XP in available to the GM per encounter, it fails to account for the effect of the *number* of enemies. A fight versus a single monster worth 1000 XP is a very different proposition than a fight against four creatures each worth 250 XP even though the total XP is the same.

The DMG accounts for this by providing XP multipliers based on the number of monsters and the number of players. The same total monster XP is multiplied by a larger value for more monsters facing a smaller party than it would be for fewer monsters facing a larger party.

So, if you are using the DMG to balance an encounter you have to total up the XP of the monsters in an encounter and then go to a different part of the DMG where you can multiply that value by the multiplier in the relevant row of a second table which would then tell you the "actual" XP of the creatures you have selected. If you had too many or too few monsters you'd have to repeat this process iteratively until you identified the correct amount of "raw" XP to yield your desired "realized" XP in an encounter. Cumbersome, right?

`dndR` provides an alternative to using the multiplier table in `xp_cost`. `xp_cost` requires the "raw" XP of the monsters you have selected, the number of monsters, and the number of player characters (PCs) and returns the "realized" XP. You can then quickly compare this with the value return by `xp_pool` to determine whether you need to add or remove creatures from your encounter.

```{r xp_cost_fxn}
dndR::xp_cost(monster_xp = 1000, monster_count = 2, party_size = 3)
```

### Quick XP Demonstration

Let's say I am running a game for four players, all level 3, and I want to design a hard encounter for them and I want to use `dndR` to balance this encounter.

To begin, I'd identify the total XP I can spend to make an encounter this difficult.

```{r xp_demonstration_1}
dndR::xp_pool(party_level = 3, party_size = 4, difficulty = 'hard')
```

Now that I know my XP threshold, I can check the value of two monsters worth (total) 500 XP against that threshold.

```{r xp_demonstration_2}
dndR::xp_cost(monster_xp = 500, monster_count = 2, party_size = 4)
```

I can see that I'm well under the XP threshold I have to play with so I can add a monster and see where that leaves me.

```{r xp_demonstration_3}
dndR::xp_cost(monster_xp = 750, monster_count = 3, party_size = 4)
```

A little over because of the multiplier, so let's say I change my mind and keep three monsters but change their identity to one with a lower XP value.

```{r xp_demonstration_4}
dndR::xp_cost(monster_xp = 600, monster_count = 3, party_size = 4)
```

Basically right on target! I can now pick out my three monsters that total up to 600 XP raw and know that they will likely* make a hard encounter for my players! (* "Likely" because there is dice rolling involved and it is possible that the monsters roll well while my players roll badly or vice versa).

## Assessing Party Abilities

It can be useful as a GM to know what abilities your party is--as a whole--strong in versus weaker in. `party_diagram` allows GMs to visualize the ability scores of every player in a party either grouped by player or by ability score. The function supports both interactive (abilities entered via the Console) and non-interactive (abilities given as a list) entries. Thank you to [Tim Schatto-Eckrodt](https://kudusch.de/) for contributing this function!

Due to the static nature of a vignette, we'll use the non-interactive path by assembling the party score list and then invoking `party_diagram`.

```{r party_diagram_1, fig.height = 6, fig.width = 6, fig.align = "center"}
# Create named list of PCs and their scores
party_list <- list(Vax = list(STR = "10", DEX = "13", CON = "14", 
                              INT = "15", WIS = "16", CHA = "12"),
                   Beldra = list(STR = "20", DEX = "15", CON = "10",
                                 INT = "10", WIS = "11", CHA = "12"),
                   Rook = list(STR = "10", DEX = "10", CON = "18", 
                               INT = "9", WIS = "11", CHA = "16"))

# Create a party diagram using that list (by player)
dndR::party_diagram(by = "player", pc_stats = party_list, quiet = TRUE)
```

You can also group the diagram by score if that is of interest.

```{r party_diagram_2, fig.height = 6, fig.width = 6, fig.align = "center"}
dndR::party_diagram(by = "ability", pc_stats = party_list, quiet = TRUE)
```

## Creating Non-Player Characters

Non-player characters (NPCs) are a key component of D&D. It can be hard to come up with NPCs on the fly that aren't directly linked to the current story arc but doing so is a great way to add realism to your world. The `npc_creator` function helps GMs speedily create some quick baseline information for a number of NPCs.

Simply specify the number of NPCs you want to generate and this function will randomly select a race and job for each NPC. You can re-run the function if the race-job combinations returned by the function don't match your current campaign tone / history.

```{r npc_creator}
dndR::npc_creator(npc_count = 2)
```

## Creating Monsters

Creatures that you create can be a great way to add flavor to an encounter or can even form the centerpiece of a larger campaign arc! The `monster_stats` and `monster_creator` functions can help GMs to quickly stat out their homebrewed monsters.

### Finding Official Monster Statistics with `monster_stats`

The DMG provides a table (see p. 274) that gives the vital statistics of creatures based on their Challenge Rating (CR) but this table can be cumbersome to compare to Experience Points (you know, the things used to determine how hard an encounter will be for your party?). `monster_stats` streamlines this process by allowing you to input either the XP you want to spend on this creature (you can use the value returned by `xp_cost`) *or* the Challenge Rating (CR) if you know it. Once either XP or CR is provided, `monster_stats` returns the creature's statistics as they appear in the DMG for a *single* creature of that difficulty.

```{r monster_stats_fxn1}
dndR::monster_stats(xp = 1000, cr = 4)
```

Challenge Rating is more than a little esoteric so feel free to ignore that argument entirely if XP is more comfortable for you!

```{r monster_stats_fxn2}
dndR::monster_stats(xp = 8000)
```

### Homebrewing Custom Monsters with `monster_creator`

If you'd rather take a more customized approach, you can use `monster_creator` instead of `monster_stats`. This function follows the advice of [Zee Bashew](https://twitter.com/Zeebashew) on how to build interesting, challenging monsters for your party. These monsters are built somewhat according to the Dungeon Master's Guide for creating monsters, partly Zee's [YouTube video on homebrewing monsters based on the video game *The Witcher*](https://www.youtube.com/watch?v=GhjkPv4qo5w), and partly on my own intuition about scaling the difficulty of a creature. Creatures are spawned randomly so you may need to re-run the function several times (or mentally modify one or more parts of the output) to get a monster that fits your campaign and players. Each creature is provided with up to five damage resistances, up to two damage immunities, and a single vulnerability. This combination allows you to build complex and mysterious homebrew monsters with plenty of opportunities for the party to have to investigate around to discover the monster's strengths and weaknesses before the final showdown.

```{r monster_creator}
dndR::monster_creator(party_level = 5, party_size = 4)
```

Note that if you use `monster_creator` you may need to help your players identify the creature's immunities and vulnerabilities *before* the actual confrontation with the creature to avoid sending them into a fight that is more difficult than your party can handle.

## `dndR` versus DMG Comparisons

See below for some comparisons between my functions and the Dungeon Master's Guide statistics they recapitulate.

### `xp_pool` vs. DMG

The DMG specifies the XP threshold *per player* for a given difficulty while my function asks for the *average* player level and the party size. This difference keeps the function streamlined and flexible for parties of any size.

Rather than embedding the DMG's table for encounter XP, `xp_pool` actually uses the formula for the line defining the XP-party level curve implicit in the DMG. This has the added benefit of being able to handle non-integer values for average party_level. 

Below is a comparison of the DMG's XP-to-party level curve versus the one obtained by `xp_pool`.

```{r xp-pool-plot, echo = F, warning = F, error = F, fig.height = 4, fig.width = 4, fig.align = "center"}
xp_df <- data.frame('pc_level' = 1:20,
                    'easy_xp' = c(25, 50, 75, 125, 250, 300, 350, 450, 550, 600, 800, 1000, 1100, 1250, 1400, 1600, 2000, 2100, 2400, 2800)
                    ) %>%
  # For each row...
  dplyr::rowwise() %>%
  # Calculate the XP for that party level using my function
  dplyr::mutate(calc_xp = dndR::xp_pool(party_level = pc_level,
                                        party_size = 1,
                                        difficulty = 'easy')) %>%
  # Rename a column more intuitively
  dplyr::rename(book_xp = easy_xp) %>%
  # Pivot longer for ease of plotting
  tidyr::pivot_longer(cols = -pc_level, 
                      names_to = 'calc_method',
                      values_to = 'xp') %>%
  # Change the entries of the calc_method column
  dplyr::mutate(
    authority = base::ifelse(test = calc_method == "book_xp",
                             yes = "DMG",
                             no = "dndR"))
# Create the plot
ggplot2::ggplot(xp_df, aes(x = pc_level, y = xp, shape = authority))  +
  geom_point(aes(fill = authority), color = 'black', size = 3,
             pch = rep(x = c(21, 24), times = 20),
             position = position_dodge(width = 0.5)) +
  geom_smooth(aes(color = authority), method = 'loess',
              formula = 'y ~ x', se = FALSE) +
  labs(x = "Party Level", y = "Experience Points (XP)") +
  scale_color_manual(values = c("#f46d43", "#74add1")) +
  scale_fill_manual(values = c("#f46d43", "#74add1")) +
  theme_classic() +
  theme(legend.position = c(0.15, 0.9),
        legend.title = element_blank(),
        axis.text.y = element_text(angle = 90, hjust = 0.5),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16))
```

### `cr_convert` vs. DMG

`cr_convert` is embedded in the `monster_stats` function and is what allows that function to handle both CR and XP inputs. The DMG specifies the XP value of a monster of any CR from 0 to 30 so `cr_convert` uses the formula of that line to avoid querying the table for this conversion.

Below is the comparison of the DMG's XP-to-CR curve and the one produced by `cr_convert`.

```{r cr-convert-plot, echo = F, warning = F, error = F, fig.height = 4, fig.width = 4, fig.align = "center"}
cr_actual <- data.frame(
  "cr" = c(0, 0.125, 0.25, 0.5, 1:30),
  "dmg_xp" = c(0, 25, 50, 100, 200, 450, 700, 1100, 1800, 2300, 2900,
               3900, 5000, 5900, 7200, 8400, 10000, 11500, 13000, 15000,
               18000, 20000, 22000, 25000, 33000, 41000, 50000, 62000,
               75000, 90000, 105000, 120000, 135000, 155000) ) %>%
  # For each row...
  dplyr::rowwise() %>%
  # Calculate XP for a given CR using `cr_convert`
  dplyr::mutate("calc_xp" = dndR::cr_convert(cr = cr)) %>%
  # Pivot to long format
  tidyr::pivot_longer(cols = dplyr::contains('_xp'),
                      names_to = "authority",
                      values_to = "xp") %>%
  # And clean up the entries of the source column
  dplyr::mutate(source = base::ifelse(test = (authority == "calc_xp"),
                                      yes = "dndR", no = "DMG"))
# Creat the plot comparison
ggplot(cr_actual, aes(x = cr, y = xp, shape = source)) +
  # geom_point(size = 3, pch = rep(x = c(17, 19), times = 34),
  #            position = position_dodge(width = 0.5)) +
  geom_point(aes(fill = source), color = 'black', size = 3,
             pch = rep(x = c(21, 24), times = 34),
             position = position_dodge(width = 0.5)) +
  geom_smooth(aes(color = source), method = 'loess',
              formula = 'y ~ x', se = FALSE) +
  # geom_smooth(formula = 'y ~ x', method = 'loess', se = F) +
  labs(x = "Challenge Rating (CR)", y = "Experience Points (XP)") +
  scale_color_manual(values = c("#c51b7d", "#4d9221")) +
  scale_fill_manual(values = c("#c51b7d", "#4d9221")) +
  theme_classic() +
  theme(legend.position = c(0.2, 0.9),
        legend.title = element_blank(),
        # axis.text.y = element_text(angle = 90, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 14),
        axis.title = element_text(size = 16))
```
