---
title: "Encounter Balancing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Encounter Balancing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r libs, echo = F, message = F}
# devtools::install_github("njlyon0/dndR", force = TRUE)
library(dndR)
```

When I am the Dungeon/Game Master (DM / GM) I find encounter balancing to be really difficult, in part because of the nest of interrelated tables and experience point multipliers outlined in the Dungeon Master's Guide (DMG) that must be consulted for novice GMs such as myself. To help other newbies, `dndR` includes two core functions aimed at simplifying encounter balancing.

### Experience Point (XP) Thresholds

The difficulty of an encounter in D&D is affected by three things:

1. Level of characters
2. Number of characters in the party
3. How difficult the GM wants to make things for their players

The DMG handles this by providing experience point (XP) thresholds based on these three factors. All enemies are worth a pre-determined amount of XP so encounters are balanced by the DMG listing the total XP of all monsters in a given fight for every level of players, party size, and difficulty. That table is useful but a little dense to work through as you're prepping potentially multiple encounters per session, so this is where `xp_pool` becomes useful.

`xp_pool` returns the amount of XP the GM can 'spend' on monsters in a given encounter to ensure the difficulty is as desired based on the three factors identified above.

```{r xp_pool}
dndR::xp_pool(party_level = 2, party_size = 4, difficulty = "medium")
```

In the above case, the DM can have their four-person, level two party face creatures worth a total of 625 XP and be reasonably confident that the difficulty will be somewhere in the middle.

### XP Multipliers

However, while knowing the amount of XP available to the DM per encounter is helpful, it fails to account for the effect of the *number* of enemies. A fight versus a single monster worth 1000 XP is a very different proposition than a fight against four creatures each worth 250 XP even though the total XP is the same.

The DMG accounts for this by providing XP multipliers based on the number of monsters and the number of players. The same total monster XP is multiplied by a larger value for more monsters facing a smaller party than it would be for fewer monsters facing a larger party.

So, if you are using the DMG to balance an encounter you have to total up the XP of the monsters in an encounter and then go to a _different part of the DMG_ where you can multiply that value by the multiplier in the relevant row of a second table which would then tell you the "actual" XP of the creatures you have selected. If you had too many or too few monsters you'd have to repeat this process iteratively until you identified the correct amount of "raw" XP to yield your desired "realized" XP in an encounter. Cumbersome, right?

The `xp_cost` function provides an alternative to using the multiplier table. `xp_cost` requires the "raw" XP of the monsters you have selected, the number of monsters, and the number of player characters (PCs) and returns the "realized" XP. You can then quickly compare this with the value return by `xp_pool` to determine whether you need to add or remove creatures from your encounter.

```{r xp_cost}
# Identify 'realized' XP of two monsters worth a total of 1,000 XP versus a 3-player party
dndR::xp_cost(monster_xp = 1000, monster_count = 2, party_size = 3)
```

## Balancing Demonstration

That was a lot of content so let's run through a quick example to see how `xp_pool` and `xp_cost` work together to make encounter balancing simpler. Let's say I am running a game for four players, all level 3, and I want to design a hard encounter for them.

To begin, I'd identify the total XP I can spend to make an encounter of this difficulty using `xp_pool`.

```{r balance-demo-1}
dndR::xp_pool(party_level = 3, party_size = 4, difficulty = 'hard')
```

Okay, so the pool of XP available for me to spend is 1,112. Keeping that threshold in mind, I can use `xp_cost` to check the value of two monsters worth (total) 500 XP against that threshold.

```{r balance-demo-2}
dndR::xp_cost(monster_xp = 500, monster_count = 2, party_size = 4)
```

I can see that I'm well under the XP threshold I have to play with so I can add a monster and see where that leaves me.

```{r balance-demo-3}
dndR::xp_cost(monster_xp = 750, monster_count = 3, party_size = 4)
```

A little over because of the multiplier, so let's say I change my mind and keep three monsters but change their identities to reduce their total XP cost.

```{r balance-demo-4}
dndR::xp_cost(monster_xp = 600, monster_count = 3, party_size = 4)
```

Basically right on target! I can now pick out my three monsters that total up to 600 XP raw and know that they will likely* make a hard encounter for my players! (* "Likely" because there is dice rolling involved and it is possible that the monsters roll well while my players roll badly or vice versa).
